The interactions of rational, pragmatic agents  \n lead to efficient language structure and use

### 2019-01-31 paper/paper.Rmd
update objective function -- integrate comments from NDG about emphasis on novelty of objective and our framing.

abstract: 
    Languages display a diverse set of distributional regularities such as the relation between a word's frequency and rank in a corpus, the distribution of dependency lengths, or the presence of lexical properties such as ambiguity. We discuss a framework for studying how these properties emerge from in-the-moment interactions of rational, pragmatic speakers and listeners. Our work takes notions of lexicon-level efficiency as a starting point, connecting these ideas to notions of conversational-level efficiency. To do so, we derive an objective function for measuring the communicative efficiency of linguistic systems and then examining the behavior of this objective in a series of simulations focusing on the communicative function of ambiguity in language. These simulations suggest that rational pragmatic agents will produce communicatively efficient systems and that interactions between such agents provide a framework for examining efficient properties of language more broadly.

@@

Zipf articulated what is now considered a *functionalist* approach to language science -- analyzing language structure and use in terms of efficiency. Such an approach might reframe our opening question as follows -- how does having \textbf{property x} make using language $\ell$ more or less useful for communication? This reframing provides an opportunity to study both language *structure* (as Zipf primarily did) or *use* in terms of efficiency. For example, Regier et al. [-@RegierKempKay2015a] showed that languages appear to organize semantic domains (form-meaning mappings) in maximally efficient ways. Likewise, Piantadosi et al. [-@Piantadosi2011a] argued that lexical ambiguity is an efficient property of any communication system when communication is contextualized.\par

Zipf’s original work and subsequent functionalist projects describe a varied set of linguistic properties and human behaviors. Common to many is a characterization of a fundamental effort-asymmetry underlying everyday communication. Simply put, what is “hard” for a speaker is likely different from what is “hard” for a listener. Zipf characterized this as follows -- purely from the standpoint of speaker effort, an optimal language $\ell_{speaker}^*$ would tend toward a vocabulary of just a single, low-cost word. Given such a language, the full set of potential meanings would be conveyed using only that word, i.e. $\ell_{speaker}^*$ would be fully ambiguous and all possible meanings would need to be disambiguated by a listener. From the standpoint of listener effort, an optimal language $\ell_{listener}^*$ would map all possible meanings to distinct words, removing a listener's need to disambiguate. In this example, speaker effort is related to *production cost* and listener effort to *understanding or disambiguation cost*. Clearly natural languages fall between the two extremes of $\ell_{speaker}^*$ and $\ell_{listener}^*$. Zipf proposed that the particular lexicon-level properties he observed were a result of these competing forces -- the pressure to jointly minimize speaker and listener effort.\par

But how does such optimization take place? The example given by Zipf [-@Zipf1949a], describes local, communicative interactions in terms of a \textit{reference game}. Speakers intend to refer to some object in the world $m$. They choose some utterance $u$ to transmit this intended meaning, $u \rightarrow m$. The listener attempts to reconstruct this intended meaning given the transmitted utterance, $m \rightarrow u$. Other functionalist projects have assumed this basic reference game setting [@RegierKempKay2015a; @Piantadosi2011a] and this simplification of the communicative act has proven productive in theoretical [@FerreriCancho2018a], simulation-based [@KirbyGriffithsSmith2014a] and empirical explorations [@HawkinsFrankeSmithGoodman2018a] of efficient language structure and use.\par

Zipf's conception of speaker and listener effort may be connected to speakers' pragmatic reasoning in the moment. Under a Gricean treatment of pragmatics, speakers and listeners are assumed to follow a set of conversational maxims in which they cooperate to transfer information (Grice, 1975). These maxims appear to emerge from efficiency concerns, however (Horn, 1984). A primary target of the current work is to formalize this connection, by providing a framework for deriving system-level efficiencies from local, interaction behavior. Our claim is that to understand an "efficient" property of a system we need to consider *how* that property is *used* efficiently. Additionally, any quantitative description of "efficiency" should be a function of both the system and users.\par

In the current work we provide a case-study for the kind of framing presented above. We choose a property of languages that could, in principle, vary freely, but shows strong regularities across languages. The explanadum is why this regularity holds. We examine lexical ambiguity as our property, extending ideas by Piantadosi et al. [-@Piantadosi2011a]. We define a novel measure of efficiency that depends on the interactional behavior of interlocutors. For our model of speakers and listeners we adopt the Rational Speech Act (RSA) framework, which is nicely attested in experimental data on pragmatic interaction. With these basic ingredients we have a measure of efficiency given language and speaker-listener agents. Finally, we show that the property of interest is prevalent in languages that optimize this measure of efficiency and also how this property is used efficiently during local interaction. The primary contribution of this work is to show how the reference game frameowrk, in combination with formal models of communication, can be used to connect ideas about system-level efficiencies to in-the-moment language use.\par

@@

### Reference games
Zipf’s example of optimal speaker- and listener-languages took the form of a reference game. We adopt that formulation here, assuming these communication games as our basic unit of analysis. In this setting, speakers and listeners are aware of a set of objects $M$, which we will refer to as *meanings* and are knowledgeable about the set of possible signals $U$ (*utterances*) that can be used to refer to a given meaning (see Figure 1). Utterances may have different relative costs, operationalized via a prior over utterances $P(U)$. Similarly, meanings differ in the relative degree to which they need to be talked about, operationalized as a prior over meanings $P(M)$\footnote{The prior over meanings are analogous to the \textit{need probabilities} assumed in previous work Regier, Kemp \& Kay (2015).}. We consider a set of contexts $C$ with an associated prior $P(C)$. Contexts encode different meaning priors, formalized as separate conditional distributions over meanings e.g. $p(M|C=c_i) \neq p(M|C=c_j)$. Finally, we consider a set of communicative events $e \in E$ where $<u, m> = e$ is a tuple of utterance, meaning pairs. Note that in the case of single context $|C|=1$ we simply have a prior over meanings $p(M)$, a simplification we will use in introducing our objective function.\par

### Languages
A language $\ell$ defines the set of semantic mappings between utterance and meanings. For example, Figure 1 contains four utterances $U = \{\text{blue}, \text{green}, \text{square}, \text{circle}\} \text{ and three meanings }M = \{\text{green-square}, \text{blue-square}, \text{green-circle}\}$. The boolean matrix describes the literal semantics of the language $\ell$. Note that will consider language to "contain ambiguous items" if there is at least one utterance $u \in U$ which can apply to multiple meanings (i.e. $|[[u_i]]| > 1$)\footnote{We use double brackets $[[\dots]]$ to represent denotation}. For example, in Figure 1 both the words "blue" and "green" are ambigous so we would say that $\ell$ contains ambiguity.

### Speakers and listeners
The Rational Speech-act framework (RSA) is computational-level theory of pragmatic language use. RSA can be understood as a formalization of essential Gricean pragmatic principles -- agents reason about one another and their shared context. For this reason, we adopt RSA as our representational framework to operationalize Gricean (rational and pragmatic) speaker-listeners [@Grice1975a]. We refer the reader to Section 1 of our supplementary materials for a more detailed description of basic RSA speaker-listener definitions, and to Goodman & Frank [-@GoodmanFrank2016a] for an overview of its appliations.\par

An RSA *speaker agent* defines a conditional distribution over utterances, mapping from intended meanings $M$ to utterances $U$ using $\ell$ in a given context $c$. That is, a speaker defines $P_{speaker}(u|m; \ell)$. We will use $S(u|m; \ell)$ as short-hand throughout.  A *listener agent* defines a conditional distribution over meanings, mapping from utterances $U$ to meanings $M$ using $\ell$ in a given context $c$. We will use $L(m|u; \ell)$ as shorthand. Note that both speakers and listeners can induce joint distributions over the set of all events $E$, although, importantly, these distributions may differ:
$$P_{speaker}(u, m; \ell) = S(u|m; \ell)p(m)$$
$$P_{listener}(u, m; \ell) = L(m|u; \ell)p(u)$$

@@

Zipf (1949) proposed that the particular distributional properties found in natural language emerge from competing speaker and listener pressures. We operationalize this objective in equation (1) -- the efficiency of a linguistic system $\ell$ being used by speaker and listener agents $S$ and $L$ is the sum of the expected speaker and listener effort to communicate over all possible communicative events $E$. Note that to simplify the introduction of this objective we assume there is just a single context ($|C|=1$). In this case $p(C=c)=1$ and there is only distribution over meanings ($p(m|c) = p(m)$). We provide a more detailed description of this derivation in the supplementary materials, both with and without context $C$.\par

@@

We assume that speaker effort is the negative log probability of an utterance (surprisal) -- intuitively, the number of bits needed to encode the utterance $u$. This particular formalization of speaker-cost is general enough to accommodate a range of instantiations that might in theory be related to production difficult via articulation effort, cognitive effort related to lexical access, etc [@BennettGoodman2015a].\par

$$\text{speaker effort} = -log_2(p(u))$$

We assume listener effort is the negative log probability of the listener's conditional $-log_2(L(m|u))$. This operationalization of listener effort is intuitively related to existing work in sentence processing in which word difficulty during comprehension is proportional to surprisal [@Hale2001a; @Levy2008a].

$$\text{listener effort} = -log_2(L(m|u; \ell))$$

Importantly, we assume that events $e = <u, m>$ are sampled according the to following generative model -- some object occurs in the world with probability $p(M=m)$, the speaker attempts to refer to that object by sampling from her conditional distribution $S(u|m)$ (i.e. $e \sim p(m)S(u|m)$). From the ingredients highlighted above, it is possible to derive the following objective between the speaker and listener joint distributions (see SI 2.2 for a complete derivation).

\begin{equation}
\begin{split}
  = -\mathbb{E}_{P_{speaker}}[log_2(P_{listener})] \\
  = H_{cross}(P_{speaker}, P_{listener}; \ell)
\end{split}
\end{equation}
From an information-theoretic perspective this objective is intuitive: cross-entropy gives us a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, this is the difference between the joint distribution assumed by the speaker $P_{speaker}$ and listener $P_{listener}$. In other words, an "efficient" language $\ell$ minimizes the distance between what speakers and listeners think.

@@

The task of understanding language is marked by a frequent need to handle various forms of ambiguity; lexical, syntactic, among others [@WasowPerforsBeaver2005a]. The ubiquity of this property, however, has been argued to provide evidence that langauges are not been optimized for communication\footnote{Chomsky (2002) famously claimed "If you want to make sure that we never misunderstand one another, for that purpose language is not well designed, because you have such properties as ambiguity." Note that this analysis is analogous to the Zipfian optimal Listener language $\ell_{listener}^*$ we described earlier and define in terms of equation 2 in SI 2.2 .} [@Chomsky2002a].

@@

We provide a toy example of their argument here. Say we have two objects in the world $m_1$ and $m_2$, and two languages $\ell_1$ and $\ell_2$. In language $\ell_1$, the low-cost utterance $u_1$ can be used to refer to both $m_1$ and $m_2$ (i.e. $[[u_1]]_{\ell_1} = \{m_1, m_2 \}$ and $[[u_2]]_{\ell_1} = \emptyset$). With language $\ell_2$, however, the low cost utterance $u_1$ can only refer to $m_1$ and a higher-cost utterance $u_2$ can refer to $m_2$ (i.e. $[[u_1]]_{\ell_2} = \{m_1\}$ and $[[u_2]]_{\ell_2} = \{m_2 \}$). While it is cheaper for a speaker to use $\ell_1$ (because she can always use a low-cost utterance), it is more difficult for a listener (because $u_1$ is ambiguous). If context isn't disambiguating then we might prefer $\ell_2$ to $\ell_1$. But, if context is disambiguating then the speaker can use $u_1$ to refer to either $m_1$ or $m_2$ and $\ell_1$ should be preferred.\par

@@

```{r plot-optimal-langs, fig.env = "figure*", fig.pos = "h", fig.width=8, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Optimal languages are more likely to contain ambiguous items as the amount of contextual information increases. Vertical axis shows the proportion of optimal languages that contain ambiguity. Horizontal axis shows the number of contexts in each condition (1-4). Red-line represents the optimal language under our Zipfian cross-entropy objective while the blue and red lines show optimal languages under speaker- and listener-only objectives. Error bars represent 95% confidence intervals."}
img <- png::readPNG("figs/fig1.png")
grid::grid.raster(img)
```

@@

We adopt the framework above, and (following Piantadosi et al. 2011), add \textit{context} ($C$). The authors argue that mapping two meanings $m_1$ and $m_2$ to a single utterance $u_1$ is useful when they can be disambiguated *in context*. In our case we consider a context $c_i \in C$ to specify a permutation of the meaning probabilities. That is, $p(m|c_1) \neq p(m|c_2)$, when there are two contexts $|C| = \{c_1, c_2\}$.This in turn, leads to an update to our linguistic efficiency objective as we now would like to consider the average speaker-listener effort over all contextualized communicative events (see section 3 SI).\par

We proceed by generating languages with different amounts of contextual support (varying the size of $|C|$). We search the space of languages, asessing if the ones which minimize our objective contain ambiguity. If context leads to more efficient commnication, then optimal languages should be more likley to be ambiguous as the amount of context increases.\par

@@

We conduct $N=2000$ simulations. For each simulation we enumerate the set of *valid* languages in which $|U|=|M|=4$ (recall that $U$ is our set of utterances and $M$ our set of meanings). Note that a language $\ell \in L$ is "valid" so long as each possible meaning in $m \in M$ can be referred to by at least one form $u \in U$ (every column of $\ell$ has some non-zero assignment) and each form maps to at least one meaning (every column has some non-zero assignment)\footnote{For the current set of simulations we consider languages in which there are four or five true values -- in a "four" language ambiguity is not possible, but in a "five" language there are ambiguous mappings.}. For a given simulation the goal is to find the language $\ell^*$ which minimizes our cross-entropy objective and then check to see if that language contains ambiguity.\par

Recall that language efficiency is both a function of the particular semantic mappings induced by that language, the speaker and listener agents ($S$ and $L$), as well as the utterance ($P(U)$), meaning ($P(M)$), and context priors ($P(C)$). Rather than assume particular structure for our utterance, meaning, and context prior distributions, for each simulation we generate $P(U) \sim \text{Dir}(1, |U|)$, $P(M|C=c) \sim \text{Dir}(1, |M|)$ (a separete conditional distribution over meanings for each context $c$), and $P(C) \sim \text{Dir}(1, |C|)$ where $\text{Dir}(1, k)$ specifies the uniform Dirichlet distribution over a $k$-dimensional probability vector.\par

@@

Following the argument given by Piantadosi et al. [-@Piantadosi2011a], we want to assess the impact of *context* on our objective. To do so we consider four conditions with $n=500$ simulations each (that is, 500 unique sets of $\{P(U), P(M|C), P(C)\}$. Our first is a *single-context* condition ($|C|=1$) -- there is a only a single context describing $P(M)$. Our second condition contains two-contexts  ($|C| = 2$) -- we consider efficiency under both $P(M|C=c_1)$ as well as $P(M|C=c_2)$. The third and fourth condition correspond accordingly with $|C|=3$ and $|C| = 4$, respectively.\par

@@

For comparison, we also examine properties of optimal languages under two additional objectives. Zipf [-@Zipf1949a] proposed that the optimal speaker language $\ell_{speaker}^*$ would only optimize speaker effort and the optimal listener language $\ell_{listener}^*$ would only optimize listener effort. We define these objectives using the first and second half of equation 1 (see SI section 2.2. for more detail).\par

@@

Figure 1. plots the proportion of optimal languages under each objective, as a function of context condition. The red line shows that as the number of contexts increases, so does the probability that the optimal language $\ell^*_{cross}$ under our speaker-listener cross-entropy objective contains ambiguity. We also plot the proportion of optimal speaker-only $\ell^*_{speaker}$ and listener-only $\ell^*_{listener}$ languages that contain ambiguity. In line with Zipf's predictions, if languages are designed only to minimize speaker effort then optimal languages will assign lower-cost forms multiple meanings (all of the speaker-optimal languages contained ambiguity). Likewise, if languages are designed only to minimize listener effort then ambiguity should always be avoided (none of the listener-optimal languages contained ambiguity).\par

We explored the degree to which ambiguity improved efficiency during communication between rational, pragmatic agents. We showed that as contextual information increased, the optimal languages under our Cross-Entropy objective were more likely to have ambiguous lexical items. Importantly, we did not see this effect under alternative objectives which only emphasized speaker- or listener effort alone.\par

While our results indicate that ambiguity is an efficient property of contextualized language these simulations assumed that speaker-listener agents could always disambiguate items in context. That is, in our four conditions both agents had perfect knowledge of the relevant conditional distributions ($P(M|C)$). This assumption may be too strong for describing much of day-to-day communication -- we seldom begin a conversation with perfect knowledge of the current context of a conversation. We explore a case in which the listener has imperfect knowledge of context at the start of the conversation in Simulation 2.\par

@@

We shift time-scales in Simulation 2 -- instead of examining the objective landscape of a space of languages, we now consider a single language that contains ambiguity and examine how that property is used over a discourse. If context is not informative with respect to meaning early in a discourse (position $i$), but is at later positions at position (position $i+k$), then the speaker should avoid using ambigous material early $i$, only using it $k$ steps later in the conversation. In this toy example we can consider “context” as analogous to a “topic” of conversation. Intuitively, we can imagine a scenario in which a reader is beginning a newspaper article. While they may have some knowledge about the article’s topic (perhaps from the title), they may not have complete knowledge of its contents, including the persons or events involved. In this setting, using an ambiguous pronoun early (position $i$) in the article can lead to misunderstanding without sufficient context to support disambiguation. But if by position $i+k$ there is sufficient contextual information it may be efficient for the writer to use ambiguous material.\par

@@

We consider consider a single language $\ell$, which contains both ambiguous and unambiguous utterances. The ambiguous utterances are less costly than the unambiguous utterances. Crucially, we do not assume that the listener knows the particular topic ($c_{current}$) of the conversation \textit{a priori}. Rather, we assume that the listener has knowledge of the set of possible topics $C = \{c_1, \dots, c_k\}$, but \textit{does not know which one is currently being used by the speaker}. Formally, this means the listener does not have access to the correct conditional distribution over meanings $P(M|C=c_{current})$ at the start of the discourse. In this setting, a speaker minimizing production effort would like to use the low-cost material as frequently as possible. However, this will incur a disambiguation cost to the listener if they are uncertain of the current context $c_{current}$ (topic of conversation).\par

>>>>

abstract: 
    Languages display a diverse set of distributional regularities such as the relation between a word's frequency and rank in a corpus, the distribution of syntactic dependency lengths across languages, or the presence of properties such as ambiguity. We discuss a framework for studying how these system-level properties emerge from local, in-the-moment interactions of rational, pragmatic speakers and listeners. To do so, we derive a novel objective function for measuring the communicative efficiency of linguistic systems in terms of the interactions of speakers and listeners. We examine the behavior of this objective in a series of simulations focusing on the communicative function of ambiguity in language. These simulations suggest that rational pragmatic agents will produce communicatively efficient systems and that interactions between such agents provide a framework for examining efficient properties of language structure and use more broadly.

@@

Zipf articulated what is now considered a *functionalist* approach to language science -- analyzing language structure and use in terms of efficiency. Such an approach might reframe our opening question as follows: how does having property \textit{x} make using language $\ell$ more or less useful for communication? Among others, this efficiency-based framing has produced a rich set of theoretical and empirical targets exploring semantic typology [@RegierKempKay2015a], properties such as ambiguity [@Piantadosi2011a] and compositionality [@KirbyGriffithsSmith2014a], and the efficient use of reduction and redundancy in production [@LevyJaeger2007a; @GenzelCharniak2002a]. These projects typically employ information-theoretic principles to characterize efficiency in natural or simulated language corpora. We adopt a different approach in this work, deriving a measure of efficiency from first principles of \textit{efficient language use} and then showing how this can lead to communicatively efficient systems.\par

Functionalist theories commonly frame language efficiency in terms of a fundamental effort-assymetry underlying everyday communication: what is “hard” for a speaker is likely different than what is “hard” for a listener. Zipf gave the following intution for how this idea -- purely from the standpoint of speaker effort, an optimal language $\ell_{speaker}^*$ would tend toward a vocabulary of a single, low-cost word. Given such a language, the full set of potential meanings would be conveyed using only that word, i.e. $\ell_{speaker}^*$ would be fully ambiguous and all possible meanings would need to be disambiguated by a listener. From the standpoint of listener effort, an optimal language $\ell_{listener}^*$ would map all possible meanings to distinct words, removing a listener's need to disambiguate. In this example, speaker effort is related to *production cost* and listener effort to *understanding or disambiguation cost*. Clearly, natural languages fall between the two extremes of $\ell_{speaker}^*$ and $\ell_{listener}^*$. Zipf proposed that the particular lexicon-level properties he observed were a result of these competing forces -- the pressure to jointly minimize speaker and listener effort.\par

But how does this optimization take place? The example given by Zipf [-@Zipf1949a], describes local, communicative interactions in terms of a \textit{reference game}. Speakers intend to refer to some object in the world $m$. They choose some utterance $u$ to transmit this intended meaning, $u \rightarrow m$. The listener attempts to reconstruct this intended meaning given the transmitted utterance, $m \rightarrow u$. Other projects have assumed this basic reference game setting [@RegierKempKay2015a; @Piantadosi2011a] and this simplification of the communicative act has proven productive in theoretical [@FerreriCancho2018a], simulation-based [@KirbyGriffithsSmith2014a] and empirical explorations [@HawkinsFrankeSmithGoodman2018a] of efficient language structure and use.\par

Zipf's conception of speaker and listener effort may be connected to speakers' pragmatic reasoning in the moment. Under a Gricean treatment of pragmatics, speakers and listeners follow a set of conversational maxims in which they cooperate to transfer information [@Grice1975a]. These maxims appear to emerge from efficiency concerns, however (Horn, 1984). We formalize this connection -- showing how system-level efficiencies can emerge from local interaction behavior. Intuitively, our claim is that to understand an "efficient" property of a system it is essential that we consider how that property is *used* efficiently. It follows that any quantitative measure of "efficiency" must be a function of both the system and its users.\par

In the current work we provide a case-study for the framing presented above. We choose a property of languages that could, in principle, vary freely, but shows strong regularities across languages. The explanadum is why this regularity holds. We examine ambiguity as our property, extending ideas by Piantadosi et al. [-@Piantadosi2011a]. We define a novel measure of efficiency that depends on the interactional behavior of speaker and listener agents. We adopt the reference game as our primary unit of interaction and model language users using the Rational Speech Act (RSA) framework -- a computional model of language use, which is nicely attested by experimental data on interaction. Using these basic ingredients, we show that the property of interest (ambiguity) is prevalent in languages that optimize our measure of efficiency. Further, we show how ambiguity is \textit{used} efficiently during local, in-the-moment interactions. The contributions of this work are twofold -- we derive a novel objective for measuring linguistic efficiency and also show how the reference game framework, in combination with formal models of communication, can be used to connect ideas about system-level efficiencies to in-the-moment language use.\par

@@

### Reference games
Zipf’s example of optimal speaker- and listener-languages took the form of a reference game. We adopt that formulation here, assuming these communication games as our basic unit of analysis. In this framework, speakers and listeners are aware of a set of objects $M$ (*meanings*) and are knowledgeable about the set of possible signals $U$ (*utterances*) that can be used to refer to a given meaning (see Figure 1). Utterances may have different relative costs, operationalized via a prior over utterances $P(U)$. Similarly, meanings differ in the relative degree to which they need to be talked about, operationalized as a prior over meanings $P(M)$\footnote{The prior over meanings are analogous to the \textit{need probabilities} assumed in previous work (Regier, Kemp \& Kay (2015).}. We consider a set of contexts $c\in C$ with an associated prior $P(C)$. Each context $c_i$ describes a different distribution over meanings e.g. $p(M|C=c_i) \neq p(M|C=c_j)$. Finally, we consider a set of communicative events $e \in E$ where $<u, m, c> = e$ is an utterance, meaning and context triple.\par

### Languages
A language $\ell$ defines the set of semantic mappings between utterance and meanings. For example, Figure 1 contains four utterances $U = \{\text{blue}, \text{green}, \text{square}, \text{circle}\} \text{ and three meanings }M = \{\text{green-square}, \text{blue-square}, \text{green-circle}\}$. The boolean matrix describes the literal semantics of the language. We define a language as "ambiguous" if there is some utterance $u \in U$ which can apply to multiple meanings (i.e. $|[[u_i]]| > 1$)\footnote{We use double brackets $[[\dots]]$ to represent denotation.}. In Figure 1 both the words "square" and "green" are ambigous so we would say that $\ell$ contains ambiguity.

### Speakers and listeners
The Rational Speech Act framework (RSA) is computational-level theory of pragmatic language use, which has been well attested in experimental work examining both speaker and listener communication behavior [@FrankGoodman2012a; @GoodmanFrank2016a]. RSA is a formalization of essential Gricean pragmatic principles -- agents reason about one another and their shared context [@Grice1975a]. We adopt RSA as our representational framework to model Gricean (rational and pragmatic) speaker-listeners in the reference game setting\footnote{See SI section 1 for a more detailed description of basic RSA speaker-listener defintions and to Goodman \& Frank [-@GoodmanFrank2016a] for an overview of its applications.}.\par

An RSA *speaker agent* defines a conditional distribution over utterances, mapping from intended meanings $M$ to utterances $U$ using $\ell$ in a given context $c$. That is, a speaker defines $P_{speaker}(u|m, c; \ell)$. We will use $S(u|m, c; \ell)$ as short-hand throughout.  A *listener agent* defines a conditional distribution over meanings, mapping from utterances $U$ to meanings $M$ using $\ell$ in a given context $c$ (i.e. $L(m|u, c;\ell)$). Note that both speakers and listeners can induce joint distributions over utterace-meaning pairs, although, importantly, these distributions may differ:
$$P_{speaker}(u, m | c; \ell) = S(u|m, c; \ell)p(m|c)$$
$$P_{listener}(u, m| c; \ell) = L(m|u, c; \ell)p(u|c)$$

@@

Zipf (1949) proposed that the particular distributional properties found in natural language emerge from competing speaker and listener pressures. We operationalize this objective in equation (1) -- the efficiency of a linguistic system $\ell$ being used by speaker and listener agents $S$ and $L$ is the sum of the expected speaker and listener effort to communicate over all possible communicative events $e \in E$.\par

@@

We assume that speaker effort is related to the suprisal of an utterance in a particular context\footnote{In the current set of simulations we consider utterances costs as independent from context (ie. $p(u|c)p(c)=p(u)p(c)$).} -- intuitively, the number of bits needed to encode the utterance $u$. This particular formalization of speaker-cost is general enough to accommodate a range of cost instantiations, such as production difficult via articulation effort, cognitive effort related to lexical access, or others [@BennettGoodman2015a].\par

$$\text{speaker effort} = -log_2(p(u|c))$$

We assume listener effort is the semantic suprisal of a meaning given an utterance. This operationalization of listener effort is intuitively related to existing work in sentence processing in which word comprehension difficulty is proportional to surprisal [@Hale2001a; @Levy2008a].

$$\text{listener effort} = -log_2(L(m|u, c; \ell))$$

Importantly, we assume that events $e = <u, m, c>$ are sampled according the to following generative model -- some context occurs in the world with probability $P(C=c)$. Within this context, an object $m$ occurs with probability $p(m|c)$. The speaker attempts to refer to that object by sampling from her conditional distribution $S(u|m, c; \ell)$ (i.e. $e \sim p(c)p(m|c)S(u|m, c; \ell)$). From the ingredients described above, it is possible to derive the following objective between the speaker and listener joint distributions (see SI 2.2 for a complete derivation).

\begin{equation}
\begin{split}
  = \mathbb{E}_{c \sim P(C)}[H_{cross}(P_{speaker}, P_{listener} | c; \ell)]\\
\end{split}
\end{equation}
From an information-theoretic perspective this objective is intuitive: The Cross-Entropy (CE) gives us a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, we have an expectation over this term -- the expected difference between the distributions assumed by the speaker $P_{speaker}$ and listener $P_{listener}$ given a set of contexts $C$. In other words, an "efficient" language $\ell$ minimizes the distance between what speakers and listeners think.

@@

The task of understanding language is marked by a frequent need to handle various forms of ambiguity: lexical, syntactic, among others[@WasowPerforsBeaver2005a]. The ubiquity of this property, however, has been argued to provide evidence that langauges are not been optimized for communication\footnote{Chomsky's (2002) claim that "... language is not well designed, because you have such properties as ambiguity" is analogous to the Zipfian optimal Listener language $\ell_{listener}^*$.} [@Chomsky2002a].

@@

We re-iterate their argument here. Say we have two objects ($m_1$ and $m_2$), two utterances ($u_1$ and $u_2$), which differ in cost, and two languages ($\ell_1$ and $\ell_2$) which describe different utterance-meaning mappings. In language $\ell_1$, the low-cost $u_1$ can be used to refer to both $m_1$ and $m_2$ ($[[u_1]]_{\ell_1} = \{m_1, m_2 \}$), but the high-cost $u_2$ cannot be used at all ($[[u_2]]_{\ell_1} = \emptyset$). By contrast, in language $\ell_2$, $u_1$ can only refer to $m_1$ and $u_2$ can only refer to $m_2$ ($[[u_1]]_{\ell_2} = \{m_1\}$ and $[[u_2]]_{\ell_2} = \{m_2 \}$). While it is cheaper for a speaker to use $\ell_1$ (because speaking is always low-cost), it is more difficult for a listener (because $u_1$ is ambiguous). Crucially, however, if context is disambiguating then the speaker can use $u_1$ to refer to either $m_1$ or $m_2$ and $\ell_1$ should be preferred to $\ell_2$.\par

@@

```{r plot-optimal-langs, fig.env = "figure*", fig.pos = "h", fig.width=8, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Optimal languages are more likely to contain ambiguous items as the amount of contextual information increases. Vertical axis shows the proportion of optimal languages containing ambiguity. Horizontal axis shows the number of context-sizes (1-4). Red-line represents the optimal language under our Zipfian cross-entropy objective while the blue and red lines show optimal languages under speaker- and listener-only objectives. Error bars represent 95 percent confidence intervals."}
img <- png::readPNG("figs/fig1.png")
grid::grid.raster(img)
```

@@

We adopt the hypothesis presented by Piantadosi, formalizing their argument in our current framework -- showing that ambiguity is an efficient property under our CE objective in the reference game setting. We proceed by generating languages with different amounts of contextual support (varying the size of $|C|$). We search the space of languages, examining whether ones which minimize our objective contain ambiguity. If context leads to more efficient commnication, then optimal languages under our objective should be more likley to be ambiguous as the amount of context increases.\par

@@

We conduct $N=2000$ simulations. For each simulation we enumerate the set of *valid* languages in which $|U|=|M|=4$ (recall that $U$ is our set of utterances and $M$ our set of meanings). Note that a language $\ell \in L$ is "valid" so long as each possible meaning in $m \in M$ can be referred to by at least one form $u \in U$ (every column of $\ell$ has some non-zero assignment) and each form maps to at least one meaning (every column has some non-zero assignment). For a given simulation the goal is to find the language $\ell^*$ which minimizes our CE objective and then check to see if that language contains ambiguity.\par

We define language efficiency as a function of the particular semantic mappings induced by that language, the speaker and listener agents ($S$ and $L$), as well as the utterance ($P(U)$), meaning ($P(M)$), and context priors ($P(C)$). Rather than assume particular structure, for each simulation we generate $P(U) \sim \text{Dir}(1, |U|)$, $P(M|C=c) \sim \text{Dir}(1, |M|)$ (a separate conditional distribution over meanings for each context $c$), and $P(C) \sim \text{Dir}(1, |C|)$, where $\text{Dir}(1, k)$ specifies the uniform Dirichlet distribution over a $k$-dimensional probability vector.\par

@@

Following the argument given by Piantadosi et al. [-@Piantadosi2011a], we want to assess the impact of *context* on the presence of ambiguity in optimal languages. To do so we consider four conditions with $n=500$ simulations each (that is, 500 unique sets of $\{P(U), P(M|C), P(C)\}$. Our first is a *one-context* condition ($|C|=1$) -- only a single distribution over meanings $P(M)$. In our two-context condition ($|C| = 2$), we consider efficiency under both $P(M|C=c_1)$ as well as $P(M|C=c_2)$. Three- and four-context conditions corresponding accordingly.\par

@@

For comparison, we also examine properties of optimal languages under two additional objectives. Zipf [-@Zipf1949a] proposed that the speaker-optimal language $\ell_{speaker}^*$ would minimize speaker effort and the listener-optimal language $\ell_{listener}^*$ would minimize listener effort. We define these objectives using the first and second half of equation 1 (see SI section 2.2. for more detail).\par

@@

Figure 1. plots the proportion of optimal languages under each objective as a function of context size. The red line shows that as the number of contexts increases, so does the probability that an optimal language $\ell^*_{cross}$ contains ambiguity under our CE objective. For comparison we also plot the proportion of speaker-optimal $\ell^*_{speaker}$ (blue line) and listener-optimal $\ell^*_{listener}$ (green line) languages that contain ambiguity. In line with Zipf's predictions, if languages are designed only to minimize speaker effort then optimal languages always contain ambiguity. If languages are designed to minimize listener effort then ambiguity is always avoided.\par

In simulation 1 we explored the degree to which ambiguity was an efficient property when communication was contextualized. We showed that as contextual information increased, the optimal languages under our CE objective were more likely to have ambiguous lexical items. Importantly, we did not see this effect under alternative objectives which only emphasized speaker- or listener effort alone.\par

While our results indicate that ambiguity is an efficient property of contextualized language, these simulations assumed that agents could always disambiguate items in context. That is, in our four conditions both agents had perfect knowledge of the relevant conditional distributions ($P(M|C)$). This assumption may be too strong for describing much of day-to-day communication -- we seldom interact with others with perfect knowledge of the current context (or topic) of conversation. To explore how ambiguity may be \textit{used} efficiencly, we next explore a case in which the listener has imperfect knowledge of context at the start of the conversation, but may infer it from a discourse history in Simulation 2.\par

@@

We shift focus in Simulation 2 -- instead of examining the landscape of a space of languages under our CE objective, we consider a single language containing ambiguity, examining how that property is used over a discourse. If context is not informative with respect to meaning early in a discourse, but is at later positions, then an efficient speaking strategy would avoid ambiguous material early, saving it for when contextual information is disambiguating. In this toy example we can consider “context” as analogous to a “topic” of conversation. As an example, imagine a scenario in which a reader is beginning a magazine article. While they may have some knowledge about the article’s topic (perhaps from the title), they may not have complete knowledge of its contents, including the persons or events involved. In this setting, using an potential low-cost, but ambiguous pronoun early may lead to misunderstanding if context is sufficiently informative. But, if by a later position enough contextual information has accumulated, it may be efficient for the writer to use the ambiguous material.\par

@@

We consider consider a single language $\ell$, which contains both ambiguous and unambiguous utterances. We assume ambiguous utterances are lower cost. Crucially, we do not assume that the listener knows the particular topic ($c_{current}$) of the conversation \textit{a priori}. Rather, that the listener has knowledge of the set of possible topics $C = \{c_1, \dots, c_k\}$, but \textit{does not know which one is currently being used by the speaker}. Formally, this means the listener does not have access to the correct conditional distribution over meanings $P(M|C=c_{current})$ at the start of the discourse.\par

Over the course of a discourse $D$, the listener tries to infer both the current topic, $c_{current}$, as well as the particular meaning $m$ of a given utterance $u$. That is, we consider updated listener and speakers ($L(m, c|u,D;\ell)$, $S(u|m,c,D;\ell)$) in which both agents can track the history of previous utterances $D$. Importantly, an agent can attempt to infer the current topic of conversation $c_{current}$ using the discourse history $D$ (see SI section 3 for more detail).\par

```{r plot-optimal-use, fig.env = "figure*", fig.pos = "h", fig.width=6, fig.height=3, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "(A) shows the empirical probability that our speaker used an ambiguous utterance as a function of discourse position. (B) shows speaker effort across the three models. (C) shows the Cross-Entropy objective under our three speaker models. Error bars represent 95 percent confidence intervals."}
